---
title: "P0: co-expression modules on airway"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{P0: co-expression modules on airway}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
has_wgcna <- requireNamespace("WGCNA", quietly = TRUE)
```

# Question and data

This article runs the P0 co-expression route on the Bioconductor
[`airway`](https://bioconductor.org/packages/release/data/experiment/html/airway.html)
ExperimentData package: four cell lines, each with an untreated and a
dexamethasone-treated RNA-seq sample. Counts, `~ cell + dex`, and
`filterByExpr()` follow the
[airway QC/DE tutorial](airway-qc-de.html). Installation can use the
network; rendering after that does not.

The scientific limit is the sample size. Eight samples cannot support a
scale-free network claim or an independent preservation test. Power and
module labels here are a runnable demonstration, not a published WGCNA
analysis.

```{r install, eval=FALSE}
BiocManager::install(c("airway", "edgeR", "DESeq2", "WGCNA"))
```

```{r load-airway}
data("airway", package = "airway")
counts <- SummarizedExperiment::assay(airway, "counts")
sample_data <- as.data.frame(
  SummarizedExperiment::colData(airway),
  optional = TRUE
)
feature_data <- as.data.frame(
  SummarizedExperiment::rowData(airway),
  optional = TRUE
)[, c("gene_id", "gene_name", "symbol", "gene_biotype"), drop = FALSE]
sample_data$dex <- factor(sample_data$dex, levels = c("untrt", "trt"))
sample_data$cell <- factor(sample_data$cell)

mae <- mae_from_matrix(
  expression = counts,
  samples = sample_data,
  row_data = feature_data,
  experiment = "airway",
  assay = "counts"
)
keep <- filter_expr(mae, "airway", group = "dex")
mae_filtered <- mae_subset_features(mae, "airway", names(keep)[keep])
vst <- transform_vst(
  mae_filtered,
  "airway",
  blind = FALSE,
  design = ~ cell + dex
)
mae_vst <- mae_add_assay(mae_filtered, "airway", vst, name = "vst")
hvg <- mae_variable_features(mae_vst, "airway", "vst", top_n = 300L)
```

Filtering retains `r format(sum(keep), big.mark = ",")` /
`r format(length(keep), big.mark = ",")` genes. Module detection uses the
`r length(hvg)` most variable VST features, chosen before the network fit.

# Soft-threshold diagnostics, then WGCNA

```{r wgcna, eval = has_wgcna}
# Same VST matrix as the QC/DE tutorial, then goodSamplesGenes() inside
# coexpr_pick_power() / coexpr_wgcna(). Eight samples cannot support a
# scale-free claim; the 0.8 reference line is drawn so that failure is visible.
power_features <- mae_variable_features(mae_vst, "airway", "vst", top_n = 500L)
power <- suppressWarnings(coexpr_pick_power(
  mae_subset_features(mae_vst, "airway", power_features),
  "airway",
  assay = "vst",
  powers = c(seq_len(10L), seq(12L, 20L, by = 2L)),
  verbose = 0
))
# n = 8 does not produce a usable scale-free fit; 6 is a conventional
# signed-network starting value for the demonstration, not a selected model.
fit <- suppressWarnings(coexpr_wgcna(
  mae_vst,
  "airway",
  assay = "vst",
  power = 6,
  top_n = 300L,
  min_module_size = 12L,
  numeric_labels = FALSE,
  verbose = 0,
  maxBlockSize = 400L
))
modules <- coexpr_modules(fit)
traits <- mae_samples(mae_vst, "airway")[, c("dex", "cell"), drop = FALSE]
trait <- coexpr_module_trait(fit, traits)
membership <- coexpr_membership(fit, mae_vst, "airway", "vst")
dex_code <- stats::setNames(
  as.integer(mae_samples(mae_vst, "airway")$dex),
  rownames(mae_samples(mae_vst, "airway"))
)
gene_significance <- stats::cor(
  t(mae_pull_assay(mae_vst, "airway", "vst")[rownames(membership), names(dex_code)]),
  dex_code
)[, 1L]
hubs <- coexpr_hubs(membership, modules, n = 3L)
expr_for_tom <- t(mae_pull_assay(mae_vst, "airway", "vst")[names(modules), , drop = FALSE])
tom <- WGCNA::TOMsimilarityFromExpr(
  expr_for_tom,
  power = 6,
  networkType = "signed",
  TOMType = "signed",
  verbose = 0
)
dimnames(tom) <- list(names(modules), names(modules))
```

```{r placeholder, eval = !has_wgcna}
# Placeholder extractors on the same airway HVG IDs so the default plots
# still build when WGCNA is not installed. Not a network fit.
features <- hvg[seq_len(18L)]
modules <- stats::setNames(
  rep(c("turquoise", "blue", "grey"), each = 6L),
  features
)
power <- list(
  fitIndices = data.frame(
    Power = c(seq_len(10L), seq(12L, 20L, by = 2L)),
    SFT.R.sq = c(
      0.04, 0.07, 0.09, 0.11, 0.10, 0.08, 0.07, 0.06, 0.05, 0.04,
      0.03, 0.03, 0.02, 0.02, 0.02
    ),
    mean.k. = c(140, 95, 70, 54, 43, 35, 29, 24, 21, 18, 14, 11, 9, 7, 6)
  ),
  powerEstimate = NA_real_
)
trait <- list(
  correlation = matrix(
    c(0.96, -0.27, -1.00, -0.03, 0.14, -0.78),
    nrow = 3L,
    dimnames = list(
      c("MEturquoise", "MEblue", "MEbrown"),
      c("dex", "cell")
    )
  ),
  p_value = matrix(
    c(0.01, 0.50, 0.01, 0.90, 0.70, 0.02),
    nrow = 3L,
    dimnames = list(
      c("MEturquoise", "MEblue", "MEbrown"),
      c("dex", "cell")
    )
  )
)
membership <- matrix(
  seq(-0.7, 0.7, length.out = 54L),
  nrow = 18L,
  dimnames = list(features, c("MEturquoise", "MEblue", "MEbrown"))
)
gene_significance <- stats::setNames(as.numeric(scale(seq_len(18L))), features)
hubs <- coexpr_hubs(membership, modules, n = 2L)
tree <- stats::hclust(stats::dist(matrix(seq_len(3L * length(features)), nrow = length(features))))
tree$labels <- features
tom <- matrix(
  0.15,
  nrow = length(features),
  ncol = length(features),
  dimnames = list(features, features)
)
tom[1:6, 1:6] <- 0.65
tom[7:12, 7:12] <- 0.55
diag(tom) <- 1
```

```{r pick-module}
module_for_plot <- {
  labels <- unique(as.character(modules))
  preferred <- setdiff(labels, c("grey", "0", "ME0"))
  if (length(preferred)) preferred[[1L]] else labels[[1L]]
}
```

# Required default plots

```{r power-plot, fig.cap = "Soft-threshold diagnostics on airway VST HVGs after filterByExpr() and goodSamplesGenes(). Panel (a) is scale-free R-squared with a dashed 0.8 reference; panel (b) is mean connectivity. Eight samples do not reach a usable scale-free fit; power 6 is a conventional signed-network starting value, not a selected model.", fig.alt = "Dual-panel WGCNA diagnostics labelled (a) scale-free R-squared with a 0.8 reference line and (b) mean connectivity versus power."}
plot_coexpr_power(power) +
  labs(title = "Soft-threshold diagnostics")
```

```{r module-plot, fig.cap = "Feature counts in each WGCNA module (or the placeholder colours when WGCNA is absent).", fig.alt = "A bar chart of module sizes coloured by conventional WGCNA module colours."}
plot_coexpr_modules(modules) +
  labs(title = "Module sizes")
```

```{r trait-plot, fig.cap = "Module eigengene correlations with dexamethasone treatment and cell line. Cells show Pearson r and significance stars (* P < 0.05, ** P < 0.01, *** P < 0.001). These are the same eight samples used to build the network.", fig.alt = "A heatmap of Pearson correlations between module eigengenes and dex or cell line, with significance stars."}
plot_coexpr_trait(trait) +
  labs(title = "Module–trait correlations")
```

```{r membership-plot, fig.cap = "Module membership versus gene significance for dexamethasone on one module. Labels are descriptive of this matrix, not an external hub validation.", fig.alt = "A scatter plot of kME against the feature–dexamethasone correlation, with a linear fit."}
plot_coexpr_membership(membership, gene_significance, module = module_for_plot) +
  labs(title = paste(module_for_plot, "MM vs GS"))
```

```{r dendrogram-plot, fig.height = 5.2, fig.cap = "Gene dendrogram for the WGCNA block, with a module colour bar. Eight samples; descriptive of this fit only.", fig.alt = "A dendrogram of co-expression features with a module colour bar under the tree."}
if (has_wgcna) {
  plot_coexpr_dendrogram(fit, modules) +
    labs(title = "Module dendrogram")
} else {
  plot_coexpr_dendrogram(tree, modules) +
    labs(title = "Module dendrogram")
}
```

```{r tom-plot, fig.height = 5.2, fig.cap = "Topological overlap for the same features used in module detection, ordered by module. A display subset of the network, not an independent validation.", fig.alt = "A TOM heatmap with a module colour bar."}
tom_features <- names(modules)[seq_len(min(60L, length(modules)))]
plot_coexpr_tom(
  tom[tom_features, tom_features, drop = FALSE],
  modules = modules[tom_features]
) +
  labs(title = "Topological overlap")
```

```{r hubs}
knitr::kable(
  utils::head(hubs, 12L),
  caption = "Top intramodular features from coexpr_hubs(). IDs are Ensembl gene IDs from airway."
)
```

Module preservation needs a second, independent MAE. `coexpr_preservation()`
is the existing wrapper; this eight-sample experiment is not split into a
fake validation cohort.

# Limits

- Eight samples are too few for a scale-free topology decision. The
  dashed 0.8 line on panel (a) is a reference, not a reached threshold.
- Modules, dendrogram order, TOM, and hub ranks come from the same VST
  matrix as the trait correlations. That is description, not external
  validation.
- `airway` is human airway smooth muscle, not a tumour cohort.

```{r session-info}
sessionInfo()
```
