---
title: "P0 stub: co-expression modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P0 stub: co-expression modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
has_wgcna <- requireNamespace("WGCNA", quietly = TRUE)
```

# What this stub shows

The P0 co-expression route is:

`transform_*` → `coexpr_pick_power()` → `coexpr_wgcna()` →
`coexpr_module_trait()` / `coexpr_membership()` → optional
`coexpr_preservation()`.

This article uses `mae_simulate()`. The annotations have no biological
meaning. WGCNA calls run only when that Suggests package is installed;
the required default plots always run on constructed extractors.

# Intended chain

```{r intended-chain, eval = has_wgcna}
mae <- mae_simulate(n_features = 80L, n_samples = 16L, seed = 20260912L)
power <- coexpr_pick_power(
  mae, "rna", assay = "log_expression",
  powers = 1:6, top_n = 40L, verbose = 0
)
fit <- coexpr_wgcna(
  mae, "rna", assay = "log_expression",
  power = 4, top_n = 40L, min_module_size = 8L,
  numeric_labels = FALSE, verbose = 0
)
modules <- coexpr_modules(fit)
traits <- mae_samples(mae, "rna")[, "condition", drop = FALSE]
trait <- coexpr_module_trait(fit, traits)
membership <- coexpr_membership(fit, mae, "rna", "log_expression")
hubs <- coexpr_hubs(membership, modules, n = 5L)
```

When WGCNA is not installed, the same extractors still accept a native-shaped
list. The plots below use that constructed path so the vignette stays offline.

```{r constructed-extractors, eval = !has_wgcna}
features <- sprintf("gene%02d", 1:18)
modules <- stats::setNames(
  rep(c("turquoise", "blue", "grey"), each = 6L),
  features
)
power <- list(
  fitIndices = data.frame(
    Power = 1:6,
    SFT.R.sq = c(0.22, 0.49, 0.74, 0.86, 0.88, 0.90),
    mean.k. = c(20, 12, 7.5, 4.8, 3.2, 2.4)
  ),
  powerEstimate = 4
)
trait <- list(
  correlation = matrix(
    c(0.61, -0.44, 0.09, 0.18, 0.52, -0.29),
    nrow = 3L,
    dimnames = list(
      c("MEturquoise", "MEblue", "MEgrey"),
      c("condition", "batch")
    )
  ),
  p_value = matrix(
    c(0.01, 0.04, 0.40, 0.22, 0.03, 0.12),
    nrow = 3L,
    dimnames = list(
      c("MEturquoise", "MEblue", "MEgrey"),
      c("condition", "batch")
    )
  )
)
membership <- matrix(
  seq(-0.7, 0.7, length.out = 54L),
  nrow = 18L,
  dimnames = list(features, c("MEturquoise", "MEblue", "MEgrey"))
)
gene_significance <- stats::setNames(
  as.numeric(scale(seq_len(18L))),
  features
)
hubs <- coexpr_hubs(membership, modules, n = 2L)
```

```{r gene-significance, eval = has_wgcna}
expression <- mae_pull_assay(mae, "rna", "log_expression")
sample_data <- mae_samples(mae, "rna")
trait_code <- stats::setNames(as.integer(sample_data$condition), rownames(sample_data))
samples <- intersect(colnames(expression), names(trait_code))
gene_significance <- stats::cor(
  t(expression[rownames(membership), samples, drop = FALSE]),
  trait_code[samples]
)[, 1L]
```

```{r pick-module}
module_for_plot <- {
  labels <- unique(as.character(modules))
  preferred <- setdiff(labels, c("grey", "0", "ME0"))
  if (length(preferred)) preferred[[1L]] else labels[[1L]]
}
```

# Required default plots

```{r power-plot, fig.cap = "Soft-threshold diagnostics from pickSoftThreshold fit indices.", fig.alt = "Two panels of scale-free R-squared and mean connectivity versus candidate WGCNA powers."}
plot_coexpr_power(power) +
  labs(title = "Soft-threshold diagnostics")
```

```{r module-plot, fig.cap = "Feature counts in each constructed or fitted WGCNA module.", fig.alt = "A bar chart of module sizes coloured by conventional WGCNA module colours."}
plot_coexpr_modules(modules) +
  labs(title = "Module sizes")
```

```{r trait-plot, fig.cap = "Module eigengene correlations with sample traits.", fig.alt = "A heatmap of Pearson correlations between module eigengenes and traits, with numeric values in each cell."}
plot_coexpr_trait(trait) +
  labs(title = "Module–trait correlations")
```

```{r membership-plot, fig.cap = "Module membership versus gene significance for the blue module.", fig.alt = "A scatter plot of kME against a trait correlation, with a linear fit."}
plot_coexpr_membership(membership, gene_significance, module = module_for_plot) +
  labs(title = paste(module_for_plot, "MM vs GS"))
```

`coexpr_hubs()` returns the top intramodular features. It is an extractor,
not a network layout.

```{r hubs}
knitr::kable(hubs)
```

Optional preservation stays a two-MAE call:

```r
coexpr_preservation(
  reference, "rna", test, "rna",
  module_colors = modules,
  reference_assay = "log_expression",
  test_assay = "log_expression"
)
```

# Limits

Synthetic data cannot justify a scientific power or module-merge choice.
Preservation needs an independent cohort. This stub does not claim
external validation.

```{r session-info}
sessionInfo()
```
