---
title: "A stateless bulk transcriptomics workflow with bulkMAE"
author: "bulkMAE developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A stateless bulk transcriptomics workflow with bulkMAE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Most sections are a cookbook and remain unevaluated because they require
optional backends or external resources. The construction and local QC chunks
below are deliberately executable during vignette builds.

## 1. Data contract

`bulkMAE` uses a `MultiAssayExperiment` as the public input because it already
defines the three relationships needed by bulk studies:

1. `experiments()` stores `SummarizedExperiment` leaves;
2. `colData()` stores primary subject/sample metadata;
3. `sampleMap()` relates assay columns to primary samples, including replicated
   measurements or assay-specific names.

The wrappers call `getWithColData()` before analysis. Consequently, a backend
sees an ordinary assay and metadata table whose rows and columns are aligned.
No result is written back to the MAE.

Public functions use `<family>_<method>()` or `<family>_<operation>()` names.
Thus `de_`, `enrich_`, `score_`, `activity_`, `coexpr_`, and `surv_` each form
one discoverable autocomplete family. The complete 0.3-to-0.4 rename table is
installed as `inst/guides/naming-migration-zh.md`.

## 2. Construct an MAE

```{r construct, eval=TRUE}
library(bulkMAE)

set.seed(1)
counts <- matrix(
  rnbinom(1000, mu = 100, size = 1 / 0.2),
  nrow = 100,
  dimnames = list(
    paste0("gene", seq_len(100)),
    paste0("sample", seq_len(10))
  )
)

sample_table <- data.frame(
  condition = rep(c("control", "treated"), each = 5),
  batch = rep(c("A", "B"), 5),
  subject_id = rep(paste0("subject", seq_len(5)), 2),
  time = seq(100, 1000, length.out = 10),
  event = rep(c(0, 1), 5),
  age = seq(45, 63, by = 2),
  stage = rep(c("I", "II"), 5),
  molecular_subtype = rep(c("A", "B"), each = 5),
  risk_score = seq(-1, 1, length.out = 10),
  row.names = colnames(counts)
)

rna <- mae_create_experiment(
  assays = list(
    counts = counts,
    log_expression = log2(counts + 1)
  ),
  col_data = sample_table
)
mae <- mae_create(list(rna = rna), sample_table)

mae_validate(mae, "rna")
mae_assays(mae, "rna")
```

For Salmon, kallisto, or RSEM estimates, use `import_tximport()` to create the
leaf. Do not round abundance-scaled values and pass them to DESeq2 unless the
DESeq2/tximport documentation explicitly supports that construction.

## 3. Quality control

Use a transformed, approximately homoscedastic assay for distances and sample
correlations. Use raw counts for library depth and expression filtering.

```{r local-qc, eval=TRUE}
metrics <- qc_library(mae, "rna", assay = "counts")
pca_local <- reduce_pca(mae, "rna", assay = "log_expression", top_n = 50)

head(metrics)
head(pca_local$x[, seq_len(2), drop = FALSE])
```

```{r qc}
keep <- filter_expr(
  mae,
  "rna",
  group = "condition"
)

qc_library(mae, "rna")

vst <- transform_vst(mae, "rna", blind = TRUE)
vst_mae <- mae_create(
  list(
    rna = mae_create_experiment(
      assays = list(vst = SummarizedExperiment::assay(vst)),
      col_data = sample_table
    )
  ),
  sample_table
)

pca <- reduce_pca(vst_mae, "rna", "vst", top_n = 50)
qc_outliers(pca)
```

`qc_outliers()` is only a review aid. It must not be used as an automatic
sample-removal rule. Investigate library metrics, metadata, and experimental
records before excluding a sample or relabelling it.

## 4. Differential expression

### DESeq2

```{r deseq2}
dds <- de_deseq2(
  mae,
  "rna",
  design = ~ batch + condition
)

de <- de_deseq2_results(
  dds,
  contrast = c("condition", "treated", "control")
)
```

### edgeR quasi-likelihood

```{r edger}
edge_test <- de_edger(
  mae,
  "rna",
  formula = ~ batch + condition,
  coef = "conditiontreated"
)
```

### limma-voom

```{r limma}
design <- model.matrix(~ batch + condition, sample_table)
contrast <- limma::makeContrasts(conditiontreated, levels = design)

limma_fit <- de_voom(
  mae,
  "rna",
  formula = ~ batch + condition,
  contrasts = contrast
)
```

### Repeated measurements

```{r dream}
dream_fit <- de_dream(
  mae,
  "rna",
  formula = ~ condition + (1 | subject_id)
)
```

Time, dose, interaction, and paired designs are model-formula problems. The
wrapper does not convert strings such as “paired” into hidden design matrices;
write the intended formula and inspect its coefficients.

## 5. Batch and unwanted variation

ComBat and ComBat-seq produce corrected matrices for visualization or methods
that explicitly need corrected input. For differential expression, modelling a
known batch in the design is usually more transparent than replacing counts.

```{r batch}
combat <- adjust_combat(
  vst_mae,
  "rna",
  assay = "vst",
  batch = "batch",
  preserve = ~ condition
)

sv <- adjust_sva(
  vst_mae,
  "rna",
  assay = "vst",
  full = ~ batch + condition,
  null = ~ batch
)
```

Surrogate variables and RUV factors must be added explicitly to the downstream
design. `bulkMAE` does not silently refit a model.

## 6. Functional interpretation

```{r enrichment}
ranks <- setNames(de$stat, rownames(de))
ranks <- ranks[is.finite(ranks)]

fgsea_result <- enrich_fgsea(ranks, hallmark_sets)
go_result <- enrich_go(
  genes = significant_entrez_ids,
  method = "ora",
  org_db = org.Hs.eg.db,
  universe = tested_entrez_ids
)

goseq_result <- enrich_goseq(
  selected = setNames(rownames(de) %in% significant_gene_ids, rownames(de)),
  genome = "hg38",
  id = "ensGene",
  bias = tested_gene_lengths,
  plot_fit = FALSE
)

gsva_scores <- score_gsva(
  vst_mae,
  "rna",
  gene_sets = hallmark_sets,
  assay = "vst"
)

pathway_activity <- activity_progeny(vst_mae, "rna", assay = "vst")
tf_activity <- activity_tf(vst_mae, "rna", assay = "vst")
```

The gene universe for ORA should be the set of genes that could have been
selected after filtering, not every gene in an annotation database.

### Choosing among gene-set tests

Use ORA for a selected gene list, preranked GSEA for a genome-wide statistic,
CAMERA/FRY/mroast when the expression matrix and design should remain in the
test, and GSVA/ssGSEA/singscore for per-sample scores.

```{r gene-set-tests}
camera_result <- enrich_camera(
  vst_mae,
  "rna",
  gene_sets = hallmark_sets,
  formula = ~ batch + condition,
  contrast = contrast[, 1],
  assay = "vst"
)

reactome_result <- enrich_reactome(
  ranks = ranks,
  method = "gsea",
  organism = "human"
)
sample_scores <- score_singscore(
  vst_mae,
  "rna",
  up_set = interferon_up,
  down_set = interferon_down,
  assay = "vst"
)
```

### decoupleR is a multi-method engine

`activity_decouple()` is not limited to transcription factors. The official engine
also exposes enrichment-style algorithms and network-aware activity methods.
The network remains an explicit input so its direction and weights are visible.

```{r decouple}
activity_methods()

resource <- activity_resource("PROGENy", organism = "human")
activities <- activity_decouple(
  vst_mae,
  "rna",
  network = resource,
  assay = "vst",
  statistics = c("mlm", "ulm", "wsum"),
  method_args = list(NULL, NULL, NULL),
  mor = "weight",
  consensus = TRUE
)
```

Use `statistics = c("aucell", "fgsea", "gsva", "ora")` only when those
statistics match the question and input scale. Use `mlm`, `ulm`, `viper`,
`wmean`, or `wsum` when regulator-target signs or weights are part of the
scientific model. Confirm current names with `activity_methods()` because the
backend, rather than bulkMAE, owns the method registry.

## 7. Networks and subtypes

```{r systems}
nmf_fit <- cluster_nmf(mae, "rna", assay = "log_expression", rank = 2:6)

consensus_fit <- cluster_consensus(
  vst_mae,
  "rna",
  assay = "vst",
  max_k = 6,
  repetitions = 1000,
  plot = NULL
)
consensus_diagnostics <- cluster_consensus_diagnostics(
  consensus_fit,
  plot = NULL
)
classes_k3 <- cluster_consensus_classes(consensus_fit, k = 3)

wgcna_fit <- coexpr_wgcna(
  vst_mae,
  "rna",
  assay = "vst",
  power = 6,
  top_n = 5000
)

genie_weights <- network_genie3(
  vst_mae,
  "rna",
  assay = "vst",
  regulators = transcription_factors
)
genie_edges <- network_genie3_links(genie_weights, top = 1000)
```

Choose WGCNA's soft threshold from the study data; the example value is not a
universal default. Likewise, decide the number of molecular subtypes using
stability and biological/clinical validation rather than a single index.
`coexpr_wgcna()` follows `goodSamplesGenes()`: unusable samples and features are
removed automatically, a warning lists the change, and the returned native
WGCNA result retains the quality object and removed names for audit.

### Differential co-expression (DCA)

In this package DCA means differential co-expression analysis. It asks whether
gene-gene correlations differ between two groups; it is distinct from
differential expression.

```{r dca}
dca <- coexpr_differential(
  vst_mae,
  "rna",
  group = "condition",
  contrast = c("control", "treated"),
  assay = "vst",
  features = highly_variable_genes
)
```

Correlation testing scales roughly with the square of the feature count. Apply
a biologically defensible, outcome-independent feature filter before DCA and
report all correlation and FDR thresholds.

### Time courses and differential usage

`de_masigpro()` is the specialized path for multi-group polynomial time-course
or dose-response models. Its `edesign` must follow maSigPro's documented
`Time`, `Replicates`, and group-indicator layout. `dtu_diffsplice()` operates
after a native edgeR or limma exon/transcript-level fit.

```{r specialized-differential}
time_result <- de_masigpro(
  time_mae,
  "rna",
  edesign = time_design,
  assay = "counts",
  degree = 2,
  counts = TRUE
)

splice_result <- dtu_diffsplice(
  exon_fit,
  gene_id = exon_annotation$gene_id,
  feature_id = exon_annotation$exon_id
)
```

## 8. Tissue deconvolution

```{r deconvolution}
fractions <- deconv(
  tpm_mae,
  "rna",
  assay = "tpm",
  method = "quantiseq"
)

music_fit <- deconv_music(
  mae,
  "rna",
  assay = "counts",
  sc_reference = reference_sce,
  clusters = "cell_type",
  samples = "donor"
)

cibersortx_table <- deconv_cibersortx_input(tpm_mae, "rna")
```

The CIBERSORTx helper stops at input preparation because execution occurs in an
external service or container. Check data-governance requirements before any
upload. BayesPrism and MuSiC require an appropriate single-cell reference; the
MAE holds the bulk data, not the reference-specific modelling choices.

## 9. Clinical models and drug reversal

Feature selection must occur inside each resampling fold. Selecting genes on
the full cohort before cross-validation leaks outcome information.

```{r clinical}
cox_fit <- surv_cox(
  vst_mae,
  "rna",
  formula = survival::Surv(time, event) ~ age + stage
)

km_fit <- surv_km(
  vst_mae,
  "rna",
  formula = survival::Surv(time, event) ~ molecular_subtype
)

roc_fit <- surv_roc(
  vst_mae,
  "rna",
  time = "time",
  event = "event",
  marker = "risk_score",
  times = c(300, 600, 900)
)

penalized <- surv_penalized(
  vst_mae,
  "rna",
  time = "time",
  event = "event",
  assay = "vst",
  folds = 5
)

query <- drug_query(ranks, n = 150)
lincs_result <- drug_lincs(query, reference_database = lincs_path)
```

For disease reversal, prioritize negative connectivity scores. Interpret them
with cell line, dose, time, mechanism, toxicity, and external validation; a
connectivity score is not clinical evidence.

## 10. Reproducibility

Record the MAE creation code, package versions, design formulas, contrasts,
feature filters, gene identifier release, pathway database release, random
seeds, and the exact external reference used for deconvolution or LINCS. The
package stays stateless so these choices remain visible in the analysis script.
