---
title: "P0 stub: immune deconvolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P0 stub: immune deconvolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
```

# What this stub shows

The P0 immune / TME route is:

pin a reference → `deconv_reference()` → `deconv()` / `deconv_music()` /
`deconv_bayesprism()` → `deconv_fractions()` → associate with subtype or
clinical columns.

CIBERSORTx is **not** the default. `deconv_cibersortx_input()` only builds a
local mixture table; it does not upload.

This article uses constructed fractions so the plots run without
immunedeconv, MuSiC, or BayesPrism. Those backends are optional Suggests.

# Pin the reference

A real study must name the single-cell or signature source, species, tissue,
and identifier type. `deconv_reference()` only aligns a user-supplied count
matrix; it does not invent reference biology.

```r
# Example only. Replace with a tissue-matched, licensed reference.
sc_reference <- deconv_reference(
  counts = gene_by_cell_counts,
  cell_data = cell_metadata,
  cell_type = "cell_type",
  sample = "donor_id"
)
music <- deconv_music(
  mae, "rna",
  sc_reference = sc_reference,
  assay = "counts"
)
fractions <- deconv_fractions(music)
```

Signature-style methods such as quanTIseq go through `deconv()` and still
need HGNC symbols and a non-log TPM-like assay. Record the immunedeconv
method name and package version in the project lockfile.

# Constructed fractions for the default plots

```{r fractions}
mae <- mae_simulate(n_features = 40L, n_samples = 8L, seed = 20260912L)
samples <- rownames(mae_samples(mae, "rna"))
fractions <- matrix(
  c(
    0.40, 0.36, 0.22, 0.25, 0.18, 0.20, 0.30, 0.28,
    0.35, 0.38, 0.33, 0.30, 0.42, 0.40, 0.25, 0.27,
    0.25, 0.26, 0.45, 0.45, 0.40, 0.40, 0.45, 0.45
  ),
  nrow = 3L,
  byrow = TRUE,
  dimnames = list(c("B_cell", "T_cell", "Myeloid"), samples)
)
fractions <- deconv_fractions(fractions)
sample_data <- mae_samples(mae, "rna")
group <- stats::setNames(sample_data$condition, rownames(sample_data))
subtype <- stats::setNames(
  factor(rep(c("A", "B"), each = 4L), levels = c("A", "B")),
  samples
)
```

# Required default plots

```{r stacked, fig.cap = "Stacked cell-type fractions for each simulated sample.", fig.alt = "Stacked bars of B-cell, T-cell, and myeloid fractions across eight samples."}
plot_deconv_stacked(fractions) +
  labs(title = "Cell-type fractions")
```

```{r box, fig.cap = "Cell-type fractions compared by the simulated condition.", fig.alt = "Boxplots of each cell type split into control and treated groups."}
plot_deconv_box(fractions, group) +
  labs(title = "Fractions by condition")
```

```{r heatmap, fig.height = 4.6, fig.cap = "Cell-type heatmap split by a placeholder subtype label.", fig.alt = "A heatmap of cell-type fractions with columns faceted into subtypes A and B."}
plot_deconv_heatmap(fractions, column_split = subtype) +
  labs(title = "Fractions by subtype")
```

xCell output is an enrichment score, not a fraction. The stacked helper
labels the y-axis as `Estimated value` when columns do not sum to one.

# Limits

These numbers are constructed for the plot contract. They are not a
deconvolution of `mae_simulate()` and must not be cited as immune-infiltration
results. A real analysis pins the reference source and keeps CIBERSORTx
execution outside the package.

```{r session-info}
sessionInfo()
```
