---
title: "P0: immune deconvolution with mae_simulate"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{P0: immune deconvolution with mae_simulate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
has_sce <- requireNamespace("SingleCellExperiment", quietly = TRUE)
has_music <- requireNamespace("MuSiC", quietly = TRUE)
```

# Question and data

The P0 immune / TME route is: pin a reference → `deconv_reference()` →
`deconv()` / `deconv_music()` / `deconv_bayesprism()` → `deconv_fractions()`
→ associate with subtype or clinical columns.

Bioconductor [`airway`](https://bioconductor.org/packages/release/data/experiment/html/airway.html)
is the preferred ExperimentData package for the other P0 articles, but it
has **no** tissue-matched single-cell atlas. Rather than add a second
simulator, this article uses the existing in-package helper
`mae_simulate()` for bulk counts and sample metadata. Grouping columns
come from that MAE. The numbers are not immune infiltration.

CIBERSORTx is not the default. `deconv_cibersortx_input()` only prepares a
local mixture table.

```{r install, eval=FALSE}
# Optional backends. A real study pins the reference atlas and MuSiC commit.
# BiocManager::install("SingleCellExperiment")
```

```{r load-simulate}
mae <- mae_simulate(n_features = 80L, n_samples = 8L, seed = 20260912L)
sample_table <- mae_samples(mae, "rna")
bulk <- mae_pull_assay(mae, "rna", "counts")
genes <- rownames(bulk)
samples <- rownames(sample_table)
knitr::kable(
  sample_table[, c("condition", "batch"), drop = FALSE],
  caption = "mae_simulate() sample metadata used for grouping. Synthetic; not a clinical cohort."
)
```

# Pin the reference, then extract fractions

`deconv_reference()` only standardizes a user-supplied count matrix. The
tiny reference below shares `mae_simulate()` gene IDs so a later MuSiC
call can run. It is built in this article, not by a new package helper,
and it is not a biological atlas.

```{r reference, eval = has_sce}
cell_types <- c("B_cell", "T_cell", "Myeloid")
n_donors <- 4L
cells_per_type <- 3L
n_cells <- n_donors * length(cell_types) * cells_per_type
cell_names <- sprintf("cell%03d", seq_len(n_cells))
cell_type <- rep(rep(cell_types, each = cells_per_type), n_donors)
donor <- rep(
  sprintf("donor%02d", seq_len(n_donors)),
  each = length(cell_types) * cells_per_type
)
type_mean <- c(B_cell = 12, T_cell = 20, Myeloid = 35)
sc_counts <- matrix(
  stats::rnbinom(length(genes) * n_cells, mu = type_mean[cell_type], size = 8),
  nrow = length(genes),
  dimnames = list(genes, cell_names)
)
storage.mode(sc_counts) <- "integer"
cell_data <- data.frame(
  cell_type = cell_type,
  sample_id = donor,
  row.names = cell_names,
  check.names = FALSE
)
sc_reference <- deconv_reference(
  counts = sc_counts,
  cell_data = cell_data,
  cell_type = "cell_type",
  sample = "sample_id"
)
sc_reference
```

```{r music, eval = has_sce && has_music}
music <- deconv_music(
  mae,
  "rna",
  sc_reference = sc_reference,
  assay = "counts"
)
fractions <- deconv_fractions(music)
```

```{r aligned-fractions, eval = !(has_sce && has_music)}
# Stand-in matrix so the default plots run without MuSiC. Columns are the
# mae_simulate() sample names; values follow condition. Not a backend fit.
treated <- as.integer(sample_table$condition == "treated")
fractions <- rbind(
  B_cell = 0.42 - 0.18 * treated,
  T_cell = 0.36 - 0.04 * treated,
  Myeloid = 0.22 + 0.22 * treated
)
colnames(fractions) <- samples
fractions <- deconv_fractions(fractions)
```

```{r groups}
group <- stats::setNames(sample_table$condition, samples)
batch <- stats::setNames(sample_table$batch, samples)
identical(colnames(fractions), samples)
```

A real study records the atlas name, species, tissue, identifier type, and
backend version. Signature methods such as quanTIseq still go through
`deconv()` and need HGNC symbols plus a non-log TPM-like assay.

# Required default plots

```{r stacked, fig.cap = "Stacked cell-type values aligned to mae_simulate() samples. These are not a deconvolution of a biological bulk cohort.", fig.alt = "Stacked bars of B-cell, T-cell, and myeloid values across eight simulated samples."}
plot_deconv_stacked(fractions) +
  labs(title = "Cell-type values by sample")
```

```{r box, fig.cap = "Values compared by the mae_simulate() condition column.", fig.alt = "Boxplots of each cell type split into control versus treated simulated samples."}
plot_deconv_box(fractions, group) +
  labs(title = "Values by condition")
```

```{r heatmap, fig.height = 4.6, fig.cap = "Heatmap split by the mae_simulate() batch column. Fill values remain synthetic unless MuSiC ran.", fig.alt = "A heatmap of cell-type values with columns faceted by simulated batch."}
plot_deconv_heatmap(fractions, column_split = batch) +
  labs(title = "Values by batch")
```

# Limits

- `mae_simulate()` annotations and any stand-in fractions are not TME
  biology and must not be cited as infiltration.
- `airway` remains the bulk source for the other P0 articles; it is
  skipped here only because it has no single-cell reference.
- xCell scores are enrichments, not fractions; the stacked helper labels
  the y-axis `Estimated value` when columns do not sum to one.
- No CIBERSORTx upload is performed.

```{r session-info}
sessionInfo()
```
