---
title: "P0: immune deconvolution on airway"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{P0: immune deconvolution on airway}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
has_sce <- requireNamespace("SingleCellExperiment", quietly = TRUE)
has_music <- requireNamespace("MuSiC", quietly = TRUE)
```

# Question and data

The P0 immune / TME route is: pin a reference → `deconv_reference()` →
`deconv()` / `deconv_music()` / `deconv_bayesprism()` → `deconv_fractions()`
→ associate with subtype or clinical columns.

Bulk airway counts come from
[`airway`](https://bioconductor.org/packages/release/data/experiment/html/airway.html).
That package has **no** tissue-matched single-cell atlas. The reference
and the plotted fractions therefore come from `mae_deconv_toy()`, an
in-package synthetic helper. Its `source` string says so. These numbers
are not immune infiltration in airway smooth muscle.

CIBERSORTx is not the default. `deconv_cibersortx_input()` only prepares a
local mixture table.

```{r install, eval=FALSE}
BiocManager::install(c("airway", "SingleCellExperiment"))
# MuSiC is optional and GitHub-hosted; a real study pins its commit.
```

```{r load-airway}
data("airway", package = "airway")
counts <- SummarizedExperiment::assay(airway, "counts")
sample_data <- as.data.frame(
  SummarizedExperiment::colData(airway),
  optional = TRUE
)
sample_data$dex <- factor(sample_data$dex, levels = c("untrt", "trt"))
sample_data$cell <- factor(sample_data$cell)

mae <- mae_from_matrix(
  expression = counts,
  samples = sample_data,
  experiment = "airway",
  assay = "counts"
)
samples <- rownames(mae_samples(mae, "airway"))
toy <- mae_deconv_toy(
  samples = samples,
  genes = rownames(counts)[seq_len(80L)],
  seed = 20260912L
)
toy$source
```

`mae_deconv_toy()` reuses the first 80 airway Ensembl IDs only so a later
MuSiC call would share identifiers. The counts and fractions themselves
are simulated.

# Pin the reference, then extract fractions

```{r reference, eval = has_sce}
sc_reference <- deconv_reference(
  counts = toy$counts,
  cell_data = toy$cell_data,
  cell_type = "cell_type",
  sample = "sample_id"
)
sc_reference
```

```{r music, eval = has_sce && has_music}
music <- deconv_music(
  mae,
  "airway",
  sc_reference = sc_reference,
  assay = "counts"
)
fractions <- deconv_fractions(music)
```

```{r toy-fractions, eval = !(has_sce && has_music)}
# Default path: documented synthetic fractions aligned to airway samples.
# Replace with deconv_fractions(deconv_music(...)) or deconv() once a
# real, licensed reference and backend are available.
fractions <- deconv_fractions(toy$fractions)
```

```{r groups}
sample_table <- mae_samples(mae, "airway")
group <- stats::setNames(sample_table$dex, rownames(sample_table))
cell_line <- stats::setNames(sample_table$cell, rownames(sample_table))
```

A real study records the atlas name, species, tissue, identifier type, and
backend version. Signature methods such as quanTIseq still go through
`deconv()` and need HGNC symbols plus a non-log TPM-like assay.

# Required default plots

```{r stacked, fig.cap = "Stacked synthetic cell-type values for each airway sample. These are toy fractions, not a deconvolution of the airway counts.", fig.alt = "Stacked bars of synthetic B-cell, T-cell, and myeloid fractions across eight airway samples."}
plot_deconv_stacked(fractions) +
  labs(title = "Toy cell-type fractions")
```

```{r box, fig.cap = "Toy fractions compared by dexamethasone treatment.", fig.alt = "Boxplots of each synthetic cell type split into untreated and treated airway samples."}
plot_deconv_box(fractions, group) +
  labs(title = "Toy fractions by dexamethasone")
```

```{r heatmap, fig.height = 4.6, fig.cap = "Toy fraction heatmap split by cell line. Cell line is the real airway annotation; the fill values remain synthetic.", fig.alt = "A heatmap of synthetic cell-type fractions with columns faceted by airway cell line."}
plot_deconv_heatmap(fractions, column_split = cell_line) +
  labs(title = "Toy fractions by cell line")
```

# Limits

- `mae_deconv_toy()` is not a reference atlas and must not be cited as TME
  biology.
- airway smooth-muscle RNA-seq is the wrong tissue for a blood or tumour
  immune signature unless a study explicitly justifies that transfer.
- xCell scores are enrichments, not fractions; the stacked helper labels
  the y-axis `Estimated value` when columns do not sum to one.
- No CIBERSORTx upload is performed.

```{r session-info}
sessionInfo()
```
