---
title: "P0 stub: molecular subtyping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P0 stub: molecular subtyping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
has_ccp <- requireNamespace("ConsensusClusterPlus", quietly = TRUE)
```

# What this stub shows

The P0 subtyping route is:

HVG or GSVA/eigengene features → `cluster_consensus()` (primary) or
`cluster_nmf()` → characterization with existing `de_*` / immune / clinical
helpers.

`mae_variable_features()` is the public HVG name helper. Pass those IDs to
`mae_subset_features()`. GSVA scores can become a new experiment with
`mae_add_experiment()`. UMAP/t-SNE stay exploratory via `plot_embedding()`.

# Feature selection is always offline

```{r hvg}
mae <- mae_simulate(n_features = 60L, n_samples = 12L, seed = 20260912L)
hvg <- mae_variable_features(mae, "rna", "log_expression", top_n = 20L)
cluster_mae <- mae_subset_features(mae, "rna", hvg)
length(hvg)
```

# Intended clustering call

```{r consensus, eval = has_ccp}
output_directory <- tempfile("bulkmae-subtype-")
dir.create(output_directory)
fit <- cluster_consensus(
  cluster_mae,
  "rna",
  assay = "log_expression",
  max_k = 4L,
  repetitions = 20L,
  item_fraction = 0.8,
  seed = 20260912L,
  title = output_directory,
  plot = NULL
)
classes <- cluster_consensus_classes(fit, k = 2L)
```

```{r constructed-consensus, eval = !has_ccp}
samples <- rownames(mae_samples(mae, "rna"))
make_consensus <- function(seed) {
  set.seed(seed)
  value <- matrix(stats::runif(length(samples)^2, 0.25, 0.95), nrow = length(samples))
  value <- (value + t(value)) / 2
  diag(value) <- 1
  dimnames(value) <- list(samples, samples)
  value
}
fit <- vector("list", 3L)
fit[[2]] <- list(
  consensusMatrix = make_consensus(21L),
  consensusClass = stats::setNames(rep(1:2, length.out = length(samples)), samples)
)
fit[[3]] <- list(
  consensusMatrix = make_consensus(22L),
  consensusClass = stats::setNames(rep(1:3, length.out = length(samples)), samples)
)
classes <- cluster_consensus_classes(fit, k = 2L)
```

Attach labels with the existing sample-metadata helper, then characterize
with `de_*` on the original counts if a real study needs that step.

```{r attach-labels}
labeled <- mae_add_sample_data(
  cluster_mae,
  "rna",
  classes,
  name = "subtype"
)
table(mae_samples(labeled, "rna")$subtype)
```

# Required default plots

```{r consensus-heatmap, fig.height = 5.2, fig.cap = "Pairwise consensus matrix for k = 2.", fig.alt = "A square heatmap of sample-to-sample consensus values ordered by the clustering solution."}
plot_cluster_consensus(fit, k = 2L) +
  labs(title = "Consensus matrix, k = 2")
```

```{r consensus-cdf, fig.cap = "Consensus CDF for each candidate k.", fig.alt = "Step functions showing the empirical distribution of pairwise consensus values at k = 2 and k = 3."}
plot_cluster_cdf(fit) +
  labs(title = "Consensus CDF")
```

```{r subtype-sizes, fig.cap = "Subtype sample counts split by the simulated condition.", fig.alt = "Stacked bars of subtype membership split by control versus treated."}
sample_data <- mae_samples(mae, "rna")
plot_cluster_sizes(
  classes,
  group = stats::setNames(sample_data$condition, rownames(sample_data))
) +
  labs(title = "Subtype sample distribution")
```

The subtype feature heatmap reuses `plot_assay_heatmap()`:

```{r subtype-heatmap, fig.height = 5.5, fig.cap = "Row-scaled HVG heatmap split by subtype. This is a descriptive view of the same samples used to define the classes.", fig.alt = "A heatmap of the selected highly variable features, with columns faceted by subtype."}
plot_assay_heatmap(
  labeled,
  "rna",
  assay = "log_expression",
  features = hvg,
  column_split = "subtype"
) +
  labs(title = "HVG heatmap by subtype")
```

# Limits

k is a scientific choice, not something this stub selects. Classes from the
same matrix are not an independent validation cohort. Do not treat UMAP or
t-SNE as a required subtype figure.

```{r session-info}
sessionInfo()
```
