---
title: "P0: molecular subtyping on airway"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{P0: molecular subtyping on airway}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  dpi = 120,
  fig.retina = 2
)
set.seed(20260912)
library(bulkMAE)
library(ggplot2)
has_ccp <- requireNamespace("ConsensusClusterPlus", quietly = TRUE)
```

# Question and data

This article runs the P0 subtyping route on
[`airway`](https://bioconductor.org/packages/release/data/experiment/html/airway.html):
HVG features → `cluster_consensus()` → sample labels → existing DE / clinical
helpers for characterization. Data construction matches the
[airway QC/DE tutorial](airway-qc-de.html).

Eight paired samples cannot define a molecular subtype system. `k` is
shown as a diagnostic, not selected as a biological truth. UMAP and t-SNE
remain exploratory and are not required figures.

```{r install, eval=FALSE}
BiocManager::install(c("airway", "edgeR", "DESeq2", "ConsensusClusterPlus"))
```

```{r load-airway}
data("airway", package = "airway")
counts <- SummarizedExperiment::assay(airway, "counts")
sample_data <- as.data.frame(
  SummarizedExperiment::colData(airway),
  optional = TRUE
)
feature_data <- as.data.frame(
  SummarizedExperiment::rowData(airway),
  optional = TRUE
)[, c("gene_id", "gene_name", "symbol", "gene_biotype"), drop = FALSE]
sample_data$dex <- factor(sample_data$dex, levels = c("untrt", "trt"))
sample_data$cell <- factor(sample_data$cell)

mae <- mae_from_matrix(
  expression = counts,
  samples = sample_data,
  row_data = feature_data,
  experiment = "airway",
  assay = "counts"
)
keep <- filter_expr(mae, "airway", group = "dex")
mae_filtered <- mae_subset_features(mae, "airway", names(keep)[keep])
vst <- transform_vst(
  mae_filtered,
  "airway",
  blind = FALSE,
  design = ~ cell + dex
)
mae_vst <- mae_add_assay(mae_filtered, "airway", vst, name = "vst")
hvg <- mae_variable_features(mae_vst, "airway", "vst", top_n = 150L)
heatmap_features <- hvg[seq_len(20L)]
cluster_mae <- mae_subset_features(mae_vst, "airway", hvg)
```

The clustering matrix uses `r length(hvg)` HVGs. The heatmap below uses the
first `r length(heatmap_features)` of that same list, fixed before looking
at cluster labels.

# Consensus clustering

```{r consensus, eval = has_ccp}
output_directory <- tempfile("bulkmae-airway-subtype-")
dir.create(output_directory)
fit <- cluster_consensus(
  cluster_mae,
  "airway",
  assay = "vst",
  max_k = 5L,
  repetitions = 50L,
  item_fraction = 0.8,
  seed = 20260912L,
  title = output_directory,
  plot = NULL
)
classes <- cluster_consensus_classes(fit, k = 2L)
```

```{r placeholder, eval = !has_ccp}
samples <- rownames(mae_samples(cluster_mae, "airway"))
make_consensus <- function(seed) {
  set.seed(seed)
  value <- matrix(
    stats::runif(length(samples)^2, 0.25, 0.95),
    nrow = length(samples)
  )
  value <- (value + t(value)) / 2
  diag(value) <- 1
  dimnames(value) <- list(samples, samples)
  value
}
fit <- vector("list", 5L)
for (k in 2:5) {
  fit[[k]] <- list(
    consensusMatrix = make_consensus(20L + k),
    consensusClass = stats::setNames(
      rep(seq_len(k), length.out = length(samples)),
      samples
    )
  )
}
fit[[2]]$consensusClass <- stats::setNames(
  as.integer(mae_samples(cluster_mae, "airway")$dex),
  samples
)
classes <- cluster_consensus_classes(fit, k = 2L)
```

```{r attach}
labeled <- mae_add_sample_data(cluster_mae, "airway", classes, name = "subtype")
sample_table <- mae_samples(labeled, "airway")
knitr::kable(
  table(subtype = sample_table$subtype, dex = sample_table$dex),
  caption = "Consensus classes versus dexamethasone. Same samples; not a validation cohort."
)
```

GSVA / eigengene features can replace HVGs by adding a score matrix with
`mae_add_experiment()` and clustering that leaf. `cluster_nmf()` is the
alternate factorization path on a non-negative assay.

# Required default plots

```{r consensus-heatmap, fig.height = 5.2, fig.cap = "Pairwise consensus matrix for k = 2 on 150 airway HVGs, with a subtype annotation bar.", fig.alt = "A heatmap of sample-to-sample consensus values for the k = 2 solution with a class annotation bar."}
plot_cluster_consensus(fit, k = 2L, annotation = classes) +
  labs(title = "Consensus matrix, k = 2")
```

```{r consensus-cdf, fig.cap = "Consensus CDF for k = 2 to k = 5. With eight samples these curves are a diagnostic, not a k-selection proof.", fig.alt = "Step functions of pairwise consensus values from k = 2 through k = 5."}
plot_cluster_cdf(fit) +
  labs(title = "Consensus CDF")
```

```{r consensus-delta, fig.cap = "Relative delta area of the consensus CDF. Shown as a diagnostic beside PAC; k is not selected from this curve.", fig.alt = "A line plot of relative delta area versus k."}
plot_cluster_delta(fit) +
  labs(title = "Relative delta area")
```

```{r consensus-pac, fig.cap = "Proportion of ambiguous clustering (PAC; consensus in (0.1, 0.9)). Lower is more decisive. Eight samples; not a subtype discovery.", fig.alt = "A bar chart of PAC at each candidate k."}
plot_cluster_pac(fit) +
  labs(title = "Proportion of ambiguous clustering")
```

```{r subtype-sizes, fig.cap = "Subtype counts split by dexamethasone treatment.", fig.alt = "Stacked bars of consensus class membership split by untreated versus treated."}
plot_cluster_sizes(
  classes,
  group = stats::setNames(sample_table$dex, rownames(sample_table))
) +
  labs(title = "Subtype sample distribution")
```

```{r subtype-heatmap, fig.height = 5.5, fig.cap = "Row-scaled heatmap of 20 pre-declared HVGs with gene symbols, split by consensus class and annotated by dexamethasone. Descriptive of the same matrix used to define the classes.", fig.alt = "A heatmap of 20 highly variable airway genes labelled by symbol, with subtype facets and a dexamethasone annotation bar."}
feature_table <- mae_feature_data(labeled, "airway")[heatmap_features, , drop = FALSE]
symbol_labels <- as.character(feature_table$symbol)
gene_names <- as.character(feature_table$gene_name)
missing_symbol <- is.na(symbol_labels) | !nzchar(symbol_labels)
symbol_labels[missing_symbol] <- gene_names[missing_symbol]
still_missing <- is.na(symbol_labels) | !nzchar(symbol_labels)
symbol_labels[still_missing] <- heatmap_features[still_missing]
names(symbol_labels) <- heatmap_features
plot_assay_heatmap(
  labeled,
  "airway",
  assay = "vst",
  features = heatmap_features,
  column_split = "subtype",
  column_annotation = "dex",
  feature_label = symbol_labels
) +
  labs(title = "HVG heatmap by subtype")
```

# Characterization without a new API

`mae_add_sample_data()` writes the class vector onto the leaf. Downstream
`de_*()` calls can include `subtype` in a design if a study needs that
contrast; on these eight samples that contrast is the same experiment, not
an independent lock. This article stops at the cross-tab with `dex`.

# Limits

- `k` is not chosen automatically and is not a subtype discovery.
  CDF, delta area, and PAC are shown together as diagnostics.
- Classes and the feature heatmap share one matrix.
- Do not treat UMAP/t-SNE as a required subtype figure.

```{r session-info}
sessionInfo()
```
